Joint maximum-likelihood of phylogeny and ancestral states is not consistent

### 2018-09-21 appendix.tex
minor changes; updated discussion

We now consider the $i$th factor of \eqref{eq:full_likelihood}.
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i \cup \overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
\end{align*}
where $\overline{Y}$ is the complement of the site pattern--valued random variable $Y$ and has the same distribution as $Y$.
Since
\begin{align*}
    \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
\end{align*}
given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}=0$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)
\end{align*}
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
\]
in general.

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} \Pr(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)]
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Although this property should not be surprising due to symmetry, we write it out for completeness.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, t; \boldsymbol\xi)$.
Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

We begin with the branch length inconsistency.

@@

We show that solutions of this form never obtain $\hat{t} = t^*$ except in cases of zero or infinite branch length.
Given Table~\ref{tab:gen-sitepatprob}, all solutions to $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$, and the only cases when we might have $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
\end{proof}

>>>>

We now consider the $i$th factor of \eqref{eq:full_likelihood}.
As a consequence of assuming a binary symmetric model, for some $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ the mapping $\patternToSplit(\alignmentColumn_i)$ has the property
\begin{align*}
    \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \Pr((\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i) \cup (\overline{\alignmentColumnRV}=\alignmentColumn_i, \overline{\ancestralStateColumnRV}=\ancestralStateColumn_i) \mid \tau, t)\\
    &= \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) + \Pr(\overline{\alignmentColumnRV}=\alignmentColumn_i, \overline{\ancestralStateColumnRV}=\ancestralStateColumn_i \mid \tau, t)\\
    &= 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)
\end{align*}
where $\overline{\alignmentColumnRV}$ is the complement of the site pattern--valued random variable $\alignmentColumnRV$ and has the same distribution as $\alignmentColumnRV$ (similarly for $\ancestralStateColumnRV$).
Since
\begin{align*}
    2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t),
\end{align*}
given $(\tau, t)$, there exist sets $\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t)$ such that $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t).
\end{align*}
In other words, for the $j$th site split, $\ancestralSplitPartition_j(\tau, t)\subseteq\mathcal{P}(\ancestralSplitSet)$ is the set of most likely ancestral state splits for that particular site split, topology and set of branch lengths, i.e., $\ancestralSplitPartition_j(\tau, t)$ is a set of sets of most likely internal node labels.
Here, $\xi_j$ is one of possibly many equiprobable ancestral state splits in $\ancestralSplitPartition_j(\tau, t)$.
For each $\alignmentColumn_i$, $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective as it can map values from $\alphabet^\nAncestralStateRows$ to all elements in $\mathcal{P}(\ancestralSplitSet)$.
This can be seen by using the definition of $\ancestralToSplit(\alignmentColumn_i, \cdot)$ and assuming $\alignmentColumn_{i,\nSiteRows}=0$, where in this case each of the $2^\nAncestralStateRows$ values of $\ancestralStateColumn$ correspond to each of the $2^\nAncestralStateRows$ elements of $\mathcal{P}(\{1,\ldots,\nAncestralStateRows\})$.
The same can be done for the case of $\alignmentColumn_{i,\nSiteRows}=1$, implying $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective.
From this we have
\begin{align*}
    \max_{\ancestralStateColumn_i} \ 2\cdot\Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t)\\
    &= \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)\\
    &= \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV = \xi_j \mid \tau, t)
\end{align*}
for some $j$.
Thus, each term in the likelihood can be collapsed into terms relating only to site splits and ancestral state splits, indexed by $j$, as opposed to individual observations, indexed by $i$.

\subsection*{Example}
We follow with an example computing these probabilities and likelihoods.
Consider the fixed, binary four-taxon tree $\tau_1$ in Fig.~\ref{fig:farris-fels-top}a.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& =: \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
\begin{equation*}
\begin{split}
& \mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \\
&\qquad \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
\end{split}
\end{equation*}
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
\[
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
\]
Internal node $1$ is the node connected to leaves $1$ and $3$ while internal node $2$ is connected to leaves $2$ and $4$.
The mapping from characters to splits in this case depends on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states in the same way as tip states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
\[
\Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\emptyset \mid \tau, t)\neq \Pr(\siteSplitRV=\siteSplit, \ancestralSplitRV=\{1,2\} \mid \tau, t)
\]
in general.

@@

The likelihood in \eqref{eq:profile_likelihood} can be written as
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \nonumber \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ \Pr(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \tau, t) \nonumber \\
                             &= \prod_{i=1}^{\nCols} \ \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \nonumber \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
for $\siteSplit_j\in\mathcal{P}(\siteSplitSet)$ and some $\xi_j\in\ancestralSplitPartition_j(\tau, t)$ with $1 \le j \le q$ where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ that project to site split $\siteSplit_j$.

Let
\[
L_\nCols''(\tau, t; \fullAlignment) = \prod_{j=1}^{\nSiteSplits} \ \left[\Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)\right] ^{\nCols_j(\fullAlignment)}
\]
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume $\nCols$ observations are generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{equation*}
\frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) = \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\end{equation*}
so that, in the $\nCols\rightarrow\infty$ limit,
\begin{equation}
\begin{split}
&    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment) \\
&\qquad \rightarrow \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t). \label{eq:site_pattern_profile_likelihood_mean}
\end{split}
\end{equation}

@@

%das: @EM I think this is actually the same Wald reference that Felsenstein tried to use.
%It works here for a fixed topology, and didn't work for him or your reference because I think the upper-semicontinuous condition doesn't hold for trees in general.
%Also I don't think the compactness condition holds---tree space isn't compact is it?
%However we now have a potential snag in that we are doing what Yang (1994) in his intro says is bad, i.e., doing ML for fixed, separate topologies and comparing them.
%We still have the behavior in the plots and the proofs, since they all assumed a fixed topology, and I think everything's fine with the arguments about Fels vs InvFels, but maybe you can let me know what you think?
For a fixed $\tau$, we show that $\hat{t}_\nCols \rightarrow \hat{t}$ for
\[
\hat{t}_\nCols = \arg\max_{t\in\mathcal{T}} \ \frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment)
\]
and
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\]
Using the notation in Section 5.2.1 in \citet{van1998asymptotic}, we let
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot \log \Pr(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t)
\]
so that
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires four conditions.
The first is that $\mathcal{T}$ is compact, which is obviously true.
The second is that
\[
E\left[\sup_{t\in\mathcal{T}} m_t\right] < \infty,
\]
and, since $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, this property holds.
The remaining conditions are on the maps
\[
\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)
\]
and
\[
t \mapsto m_t(\alignmentColumn).
\]
We need the first map to be measurable, which is evident since the domain $\alphabet^\nSiteRows$ of the mapping is a finite set, and so all subsets of the domain are also finite and thus measurable.
Finally, we must have the the second mapping be upper-semicontinuous for almost all $\alignmentColumn$.
For a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
If we move about in $\mathcal{T}$, a different ancestral state split becomes more likely, though when we maximize over ancestral state splits we obtain a continuous function since the maximum over continuous functions is also continuous.
This ensures the upper-semicontinuous property of this mapping, and shows $\hat{t}_\nCols \rightarrow \hat{t}$, allowing our consistency results to be proved using $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$.

@@

Consider the InvFels tree $\tau_1$ with arbitrary fidelities, i.e., $t=\{x_1,y_1,x_2,y_2,w\}$.
Next we show that the likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$ remains unchanged if $x_1$ and $x_2$ are exchanged or if $y_1$ and $y_2$ are.
Although this property should not be surprising due to symmetry, we write it out for completeness.
This holds for a general $t$, and thus holds setting $t=t^*$.
Using the Hadamard transform, we calculate the generating probabilities on the InvFels tree.
For site split $\emptyset$,
\begin{align*}
    \Pr(\siteSplitRV=\emptyset\mid \tau_1, t) & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
Similarly, for site split $\{1,3\}$,
\[
    \Pr(\siteSplitRV=\{1,3\}\mid \tau_1, t) = \frac{1}{8} (1 + x_1x_2 +  y_1y_2 -  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2),
\]
which also is invariant to exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.

@@

Thus exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t}(\tau_1, t; \boldsymbol\xi)$.
Therefore we can reduce the number of candidate likelihoods we need to search by, without loss of generality, assuming $x_2 \ge x_1$ and $y_2 \ge y_1$, with these likelihoods given in Table~\ref{tab:likelihoods} after maximizing over ancestral state splits.

@@

We begin by showing an inconsistency in branch length estimation on the InvFels tree.

@@

We show that solutions of this form never obtain $\hat{t} = t^*$ except in cases of zero or infinite branch length.
Given Table~\ref{tab:gen-sitepatprob}, all solutions to $\hat{x}_1$ have the form
\[
\hat{x}_1 = a_{x_1,0} + a_{x_1,1}(x^*)^2 + a_{x_1,2}(y^*)^2 + a_{x_1,3} x^*(y^*)^2 + a_{x_1,4}(x^*)^2(y^*)^2.
\]
where $a_{x_1,k}$ are constants independent of $x^*$ and $y^*$---in fact, $a_{x_1,k}$ takes values in the set $\{i/8 : i=-4,-3,\ldots,7,8\}$.
The true branch fidelity for $x_1$ is $x^*$, and the only cases to possibly obtain $\hat{x}_1=x^*$ are when $y^*=1$ or when $(x^*)^2=x^*$, i.e., one of the generating branch parameters is zero or infinite length; the same is true for $x_2$.
A similar argument for $y_1,y_2,$ and $w$ shows that estimates can only be consistent when $(y^*)^2=y^*$, i.e., $y^*=0$ or $y^*=1$.
\end{proof}

We now proceed to show there exist $x^*$ and $y^*$ such that the interior branch parameter $w$ is estimated as exactly one, indicating convergence to a multifurcating topology.
